---
title: "Unimodal functions and multi-species SDMs using mixed-effects models"
author: "Petr Keil"
date: "October 6, 2017"
output: 
  html_document: 
    highlight: pygments
    number_sections: yes
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here I attempt to clarify some ideas that we discussed with Duarte Viana today. The material will probably seem a bit random to a random reader.

# How does a parabola on a log scale look on a linear scale?

Let's consider an `Env` variable representing a gradient of an environmental condition. We can model logarithm of abundance (`log_Abund`) as a regular parabola (2nd-degree polynomial): $$\log Abund = 2Env - 4Env^2$$

```{r}
  Env <- seq(-2, 2, by = 0.1)
  log_Abund <- 2*Env - 4*(Env^2)
```

Now we can plot the abunance on both logarithmic and linear scale. We can see that
what we modelled as a parabola going all the way down to -20, is actually a bell-shaped curve on the linear scale:

```{r, fig.width=10}
  par(mfrow=c(1,2))
  plot(Env, log_Abund, type="l", main="Abundance on a log scale")
  plot(Env, exp(log_Abund), type="l", main="Abundance on a linear scale")
```

**Lesson:** Polynomial **GLMs (and related models) with log-link function** can indeed
be used to model non-linear unimodal responses of species to environment. The same logic applies for **logit link functions**, e.g. in models for binomially distributed data.

# Can the 2nd-degree polynomial fit a very narrow responses?

Is a parabola fitted on the log scale flexible enough to fit extremely narrow 
species responses to environment, where the species is missing from most of the 
gradient?

Here I manually generate a species with a very narrow niche, and with a lots of zeroes in the data:

```{r}
  Env <- seq(0, 1, by = 0.01)
  Abund <- rep(0, times = length(Env))
  Abund[c(48:52)] <- c(5, 7, 9, 5, 1)
  plot(Env, Abund, main="Only data")
```

Here I fit the parabola using `glm`; setting the family to `"poisson"` automatically invokeslog link function:

```{r}
  m1 <- glm(Abund~poly(Env,2), family="poisson")
```

And this is how the fitted model looks like:

```{r}
  plot(Env, Abund, main="Data and the model")
  lines(Env, predict(m1, type="response"))
```

**Lesson:** Even very simple polynomial GLMs can very well describe responses
of species with very narrow optima along an environmental gradients.

# Multi-species mixed-effects SDM in lme4

Here I will use the `vegan` dataset `mite` to show how a simple multi-species
SDM can be done.

```{r, message=FALSE, warning=FALSE}
  library(vegan)
  library(lme4)
  library(tidyr)
  library(ggplot2)
```

## Prepare the data

```{r}
  data(mite)
  data(mite.env)
  data(mite.xy)
  
  X <- data.frame(mite.xy, SubsDens = mite.env$SubsDens, mite)
  head(X)
```

I need to rearrange the data from the site X specie matrix to a long format
using the `gather` function from `tidyr`:

```{r}
 mite.long <- gather(data=X, species=colnames(mite), factor_key=TRUE)
 names(mite.long)[4:5] <- c("species","N")
 head(mite.long)
```

I will scale the continual predictor in the data to 0 mean and unit variance:

```{r}
  mite.long$SubsDens <- scale(mite.long$SubsDens )
```

## Fit the mixed-effect models

I will fit two models, each of them with varying slopes and varying intercepts, one with a linear response, the other with a polynomial response to the environment (substrate density `SubsDens`):

```{r}
m.lin <- glmer(N ~ SubsDens + (1 + SubsDens|species),
               data=mite.long, family="poisson")

m.poly <- glmer(N ~ poly(SubsDens,degree=2) + (1 + poly(SubsDens, degree=2)|species),
                data=mite.long, family="poisson")
```

The polynomial model obviously had problems with convergence.

## Extract model predictions

```{r, fig.width=10}
  preds <- data.frame(mite.long, preds=predict(m.poly, type="response"))
```

## Plot the predictions

### Log scale

```{r, fig.width=10}
  ggplot(data=preds, aes(x=SubsDens, y=log(N))) +
    geom_point(aes(colour=species)) + 
    geom_line( aes(x=SubsDens, y=log(preds), colour=species) )
```

### Linear scale

```{r, fig.width=10}
  ggplot(data=preds, aes(x=SubsDens, y=N)) +
    geom_point(aes(colour=species)) + 
    geom_line( aes(x=SubsDens, y=preds, colour=species) ) +
    ylim(c(0,100))
```

# Session info

```{r}
  sessionInfo()
```








